\MakeUppercase{Layered Partitions of $k$-Planar Graphs} \thanks{This work was partly funded by NSERC and MRI.}

### 2019-05-28 k-planar.tex
Uniform statements of lemmas
Small fixes

Layered $H$-partitions of small layered width in which $H$ has some additional property are a useful tool for a number of graph problems.  One such useful property is small \emph{treewidth}, which we now define. A \emph{tree-decomposition} $\mathcal{K}$ of a graph $G$ consists of a tree $K$ and a collection $\mathcal{K}=\{B_x:x\in V(K)\}$ of subsets of $V(G)$ indexed by nodes of $K$ such that (i)~for every $vw\in E(G)$, there exists some $x\in V(K)$ with that $v,w\in B_x$; and (ii)~for every $v\in V(G)$, $K[x] = K[\{y: x\in B_y\}]$ is connected.  The \emph{width} of the tree-decomposition $\mathcal{K}$ is $\max\{|B_x|:x\in V(K)\}-1$.  The \emph{treewidth} $\tw(G)$ of a graph $G$ is the minimum integer $t$ such that $G$ has a tree-decomposition of width $t$.

 \citet{dujmovic.joret.ea:planar} show that, if each graph $G$ in some family $\mathcal{G}$ of graphs has a layered $H$-partition for which the layered-width of the partition and the treewidth of $H$ are each upper-bounded by some constant $c_\mathcal{G}$, then each of the following quantities are upper bounded by a constant, for every $G\in\mathcal{G}$: queue number, track number, and layered treewidth. \citet{dujmovic.esperet.ea:planar} add non-repetitive chromatic number to this list.  \citet{dujmovic.joret.ea:planar} show that two common families of graphs have such $H$-partitions: planar graphs and (more generally) apex-minor-free graphs.\footnote{A graph $G$ is $t$-apex if it contains a set $A$ of at most $t$ vertices such that $G-A$ is planar. A graph $M$ is a \textit{minor} of a graph $G$ if a graph isomorphic to $M$ can be obtained from a subgraph of $G$ by contracting edges. \note{PM}{Help me finish this definition of apex-minor-free.}}

In this paper we show that another family of graphs admits layered $H$-partitions of bounded layered-width in which $H$ has bounded treewidth: $k$-planar graphs, which we now define. An \emph{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a closed curve\footnote{A closed curve in a surface $S$ is a continuous function $f:[0,1]\to S$. The points $f(0)$ and $f(1)$ are called the \emph{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.} in $\R^2$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior.  A \emph{crossing} in an embedded graph $G$ is a triple $(p,vw,xy)$ with $p\in\R^2$, $vw,xy\in E(G)$ and such that $p\in (vw\cap xy)\setminus\{v,w,x,y\}$. An embedded graph $G$ is \emph{$k$-plane} if each edge of $G$ takes part in at most $k$ crossings.  A (not necessarily embedded) graph $G'$ is \emph{$k$-planar} if there exists $k$-plane graph $G$ isomorphic to $G'$.

@@

We prove the following result:
\begin{thm}\thmlabel{k-planar}
  Every $k$-planar graph has a layered $H$-partition of layered-width at most $24k^2 + 60k + 36$ in which $H$ has treewidth at most $\treewidth$.
\end{thm}

@@

Prior to this work, the strongest structural characterization of $k$-planar graphs was in terms of layered treewidth, which we now define.  A \emph{layered tree-decomposition} $(\mathcal{L},\mathcal{K})$ consists of a layering $\mathcal{L}$ and a tree-decomposition $\mathcal{K}$ of $G$. The layered-width of $(\mathcal{L},\mathcal{K})$ is $\max\{|L\cap B|: L\in \mathcal{L},\, B\in \mathcal{K}\}$.  The \emph{layered treewidth} of $G$ is the minimum layered-width of any layered tree-decomposition of $G$.

@@

\begin{thm}\thmlabel{g-k-planar}
  Every $(g,k)$-planar graph has a layered $H$-partition of layered-width at most $\max\{8gk^2 + 12gk + 4g, 24k^2 + 36k + 12\}$ in which $H$ has treewidth at most $\treewidth$.

  Every $(g,1)$-planar graph has a layered $H$-partition of layered-width at most $\max\{48g, 30\}$ in which $H$ has treewidth at most $4$.  
\end{thm}

@@

% \subsubsection{Layered $H$-Partitions from PS-Trees}
% \seclabel{k-to-h}
% 
% Before continuing, it is worth pointing out how an $H$-partition of $\Delta-\{r\}$ can be obtained from a PS-tree $K=K(\Delta,T,r)$. (Note: There is nothing new in this section.  This is just an alternative presentation of material from \citet{dujmovic.joret.ea:planar}.) Given $K=K(\Delta,T,r)$, we obtain a layered $H$-partition $(\mathcal{L},\mathcal{P})$ as follows. 
% 
% We use the layering $\mathcal{L}=\{V_0,V_1,\ldots,V_h\}$ where $V_i=\{ v\in V(\Delta)\setminus\{r\}: \dist_{\Delta}(r,v)=i\}$ and $h=\max\{\dist_\Delta(r,v):v\in V(\Delta)\}$.  The fact that $T$ is a BFS tree ensures that $\mathcal{L}$ is, indeed, a layering of $\Delta-\{r\}$.  We use the vertex partition $\mathcal{P}=\{ S_x:=V(Y_x) : x\in V(K)\}$.  \propref{partition} ensures that $\mathcal{P}$ is, indeed a partition of $V(\Delta)-\{r\}$.
% 
% The resulting layered $H$-partition $(\mathcal{L},\mathcal{P})$ has layered width at most 3 because, for each $x\in V(K)$, $S_x$ is made up of vertices from at most 3 vertical paths in $T$.  These are either the three vertices of $Y_{r_k}$ (if $x=r_K$) or the vertices of the open tripod $Y_x$ whose vertices are included in three vertical paths $Q_1,Q_2,Q_3$ that form the legs of the closed tripod $\overline{Y}_x$.  Therefore $|S_x\cap V_i|\le 3$.
% 
% To show that the quotient graph $H=\Delta/\mathcal{P}$ has small treewidth, we produce a tree-decomposition $\mathcal{K}=\{B_x: x\in V(K)\}$ of $H$ using the tree $K$.  For each $x\in V(K)$, the subtree $K[x]=K[\{y:x\in B_y\}]$ contains
% $x$ as well as every node $y$ such that $V(Y_x)\cap V(F_y) \neq\emptyset$.
% It is straightforward to verify that $K[x]$ is connected for each $x\in V(K)$. The fact that $F_x$ is a cycle in $\Delta$ and \propref{ancestor-boundary} ensures that, if $xy\in E(H)$, then one of $x$ or $y$ is a $K$-ancestor of the other. \propref{ancestor-boundary} also implies that, for any node $y\in V(K)$, $|B_y|\le 4$, since $B_y$ contains $y$ and at most 3 $K$-ancestors of $y$.
% 
% Therefore $(\mathcal{L},\mathcal{P})$ is a layered $H$-partition of $\Delta-\{r\}$ with layered-width 3 in which $H$ has treewidth 3.  Furthermore, for each $P\in\mathcal{P}$, $\Delta[P]$ is connected.  Since $\Delta$ is planar, this implies that $H$ is planar.\footnote{It is well-known and easy to see that any graph $\Delta'$ obtained by contracting an edge in a planar graph $\Delta$ is also planar.  Repeatedly applying this fact implies that $\Delta/\mathcal{P}$ is planar provided that $\Delta[P]$ is connected for every $P\in\mathcal{P}$.}

@@

A $c$-colouring $\phi$ of $G$ is $p$-centered if, for every connected subgraph $X\subseteq G$, $|\{\phi(v):v\in V(X)\}| > p$ or there exists some $v\in V(X)$ such that $\phi(v)\neq \phi(w)$ for every $w\in V(X)\setminus\{v\}$.  In words, either $X$ receives more than $p$ colours or some vertex of $X$ receives a unique colour.  Let $\chi_p(G)$ denote the smallest integer $c$ such that $G$ has a $p$-centered $c$-colouring.

@@

\begin{lem}
  Every $(g,\delta)$-string graph $G$ is a $(g,2\delta^2)$-planar graph.
\end{lem}

>>>>

Layered $H$-partitions of small layered width in which $H$ has some additional property are a useful tool for a number of graph problems.  One such useful property is small \emph{treewidth}, which we now define. A \emph{tree-decomposition} $\mathcal{K}$ of a graph $G$ consists of a tree $K$ and a collection $\mathcal{K}=\{B_x:x\in V(K)\}$ of subsets of $V(G)$ indexed by nodes of $K$ such that (i)~for every $vw\in E(G)$, there exists some $x\in V(K)$ with $v,w\in B_x$; and (ii)~for every $v\in V(G)$, $K[x] = K[\{y: x\in B_y\}]$ is connected.  The \emph{width} of the tree-decomposition $\mathcal{K}$ is $\max\{|B_x|:x\in V(K)\}-1$.  The \emph{treewidth} $\tw(G)$ of a graph $G$ is the minimum integer $t$ such that $G$ has a tree-decomposition of width $t$.

 \citet{dujmovic.joret.ea:planar} show that, if each graph $G$ in some family $\mathcal{G}$ of graphs has a layered $H$-partition for which the layered width of the partition and the treewidth of $H$ are each upper-bounded by some constant $c_\mathcal{G}$, then each of the following quantities are upper bounded by a constant, for every $G\in\mathcal{G}$: queue number, track number, and layered treewidth. \citet{dujmovic.esperet.ea:planar} add non-repetitive chromatic number to this list.  \citet{dujmovic.joret.ea:planar} show that two common families of graphs have such $H$-partitions: planar graphs and (more generally) apex-minor-free graphs.\footnote{A graph $G$ is $t$-apex if it contains a set $A$ of at most $t$ vertices such that $G-A$ is planar. A graph $M$ is a \textit{minor} of a graph $G$ if a graph isomorphic to $M$ can be obtained from a subgraph of $G$ by contracting edges. \note{PM}{Help me finish this definition of apex-minor-free.}}

In this paper we show that another family of graphs admits layered $H$-partitions of bounded layered width in which $H$ has bounded treewidth: $k$-planar graphs, which we now define. An \emph{embedded graph} $G$ is a graph with $V(G)\subset\R^2$ in which each edge $vw\in E(G)$ is a closed curve\footnote{A closed curve in a surface $S$ is a continuous function $f:[0,1]\to S$. The points $f(0)$ and $f(1)$ are called the \emph{endpoints} of the curve.  When there is no danger of misunderstanding we treat a curve $f$ as the point set $\{f(t):0\le t\le 1\}$.} in $\R^2$ with endpoints $v$ and $w$ and not containing any vertex of $G$ in its interior.  A \emph{crossing} in an embedded graph $G$ is a triple $(p,vw,xy)$ with $p\in\R^2$, $vw,xy\in E(G)$ and such that $p\in (vw\cap xy)\setminus\{v,w,x,y\}$. An embedded graph $G$ is \emph{$k$-plane} if each edge of $G$ takes part in at most $k$ crossings.  A (not necessarily embedded) graph $G'$ is \emph{$k$-planar} if there exists $k$-plane graph $G$ isomorphic to $G'$.

@@

We prove the following result:
\begin{thm}\thmlabel{k-planar}
  Every $k$-planar graph has a layered $H$-partition of layered width at most $24k^2 + 60k + 36$ in which $H$ has treewidth at most $\treewidth$.
\end{thm}

@@

Prior to this work, the strongest structural characterization of $k$-planar graphs was in terms of layered treewidth, which we now define.  A \emph{layered tree-decomposition} $(\mathcal{L},\mathcal{K})$ consists of a layering $\mathcal{L}$ and a tree-decomposition $\mathcal{K}$ of $G$. The layered width of $(\mathcal{L},\mathcal{K})$ is $\max\{|L\cap B|: L\in \mathcal{L},\, B\in \mathcal{K}\}$.  The \emph{layered treewidth} of $G$ is the minimum layered width of any layered tree-decomposition of $G$.

@@

\begin{thm}\thmlabel{g-k-planar}
  Every $(g,k)$-planar graph has a layered $H$-partition of layered width at most $\max\{8gk^2 + 12gk + 4g, 24k^2 + 36k + 12\}$ in which $H$ has treewidth at most $\treewidth$.

  Every $(g,1)$-planar graph has a layered $H$-partition of layered width at most $\max\{48g, 30\}$ in which $H$ has treewidth at most $4$.  
\end{thm}

@@

% \subsubsection{Layered $H$-Partitions from PS-Trees}
% \seclabel{k-to-h}
% 
% Before continuing, it is worth pointing out how an $H$-partition of $\Delta-\{r\}$ can be obtained from a PS-tree $K=K(\Delta,T,r)$. (Note: There is nothing new in this section.  This is just an alternative presentation of material from \citet{dujmovic.joret.ea:planar}.) Given $K=K(\Delta,T,r)$, we obtain a layered $H$-partition $(\mathcal{L},\mathcal{P})$ as follows. 
% 
% We use the layering $\mathcal{L}=\{V_0,V_1,\ldots,V_h\}$ where $V_i=\{ v\in V(\Delta)\setminus\{r\}: \dist_{\Delta}(r,v)=i\}$ and $h=\max\{\dist_\Delta(r,v):v\in V(\Delta)\}$.  The fact that $T$ is a BFS tree ensures that $\mathcal{L}$ is, indeed, a layering of $\Delta-\{r\}$.  We use the vertex partition $\mathcal{P}=\{ S_x:=V(Y_x) : x\in V(K)\}$.  \propref{partition} ensures that $\mathcal{P}$ is, indeed a partition of $V(\Delta)-\{r\}$.
% 
% The resulting layered $H$-partition $(\mathcal{L},\mathcal{P})$ has layered width at most 3 because, for each $x\in V(K)$, $S_x$ is made up of vertices from at most 3 vertical paths in $T$.  These are either the three vertices of $Y_{r_k}$ (if $x=r_K$) or the vertices of the open tripod $Y_x$ whose vertices are included in three vertical paths $Q_1,Q_2,Q_3$ that form the legs of the closed tripod $\overline{Y}_x$.  Therefore $|S_x\cap V_i|\le 3$.
% 
% To show that the quotient graph $H=\Delta/\mathcal{P}$ has small treewidth, we produce a tree-decomposition $\mathcal{K}=\{B_x: x\in V(K)\}$ of $H$ using the tree $K$.  For each $x\in V(K)$, the subtree $K[x]=K[\{y:x\in B_y\}]$ contains
% $x$ as well as every node $y$ such that $V(Y_x)\cap V(F_y) \neq\emptyset$.
% It is straightforward to verify that $K[x]$ is connected for each $x\in V(K)$. The fact that $F_x$ is a cycle in $\Delta$ and \propref{ancestor-boundary} ensures that, if $xy\in E(H)$, then one of $x$ or $y$ is a $K$-ancestor of the other. \propref{ancestor-boundary} also implies that, for any node $y\in V(K)$, $|B_y|\le 4$, since $B_y$ contains $y$ and at most 3 $K$-ancestors of $y$.
% 
% Therefore $(\mathcal{L},\mathcal{P})$ is a layered $H$-partition of $\Delta-\{r\}$ with layered width 3 in which $H$ has treewidth 3.  Furthermore, for each $P\in\mathcal{P}$, $\Delta[P]$ is connected.  Since $\Delta$ is planar, this implies that $H$ is planar.\footnote{It is well-known and easy to see that any graph $\Delta'$ obtained by contracting an edge in a planar graph $\Delta$ is also planar.  Repeatedly applying this fact implies that $\Delta/\mathcal{P}$ is planar provided that $\Delta[P]$ is connected for every $P\in\mathcal{P}$.}

@@

A $c$-colouring $\phi$ of $G$ is \emph{$p$-centered} if, for every connected subgraph $X\subseteq G$, $|\{\phi(v):v\in V(X)\}| > p$ or there exists some $v\in V(X)$ such that $\phi(v)\neq \phi(w)$ for every $w\in V(X)\setminus\{v\}$.  In words, either $X$ receives more than $p$ colours or some vertex of $X$ receives a unique colour.  Let $\chi_p(G)$ denote the smallest integer $c$ such that $G$ has a $p$-centered $c$-colouring.

@@

\begin{lem}
  Every $(g,\delta)$-string graph is $(g,2\delta^2)$-planar.
\end{lem}
